Cell Count Analysis for Lapse × DR × Treatment Interaction (Monotherapy Only)

### 2025-09-04 reanalysis-5-outcome_analysis-full_interaction.qmd
full

---
title: "Cell Count Analysis for Full Four-Treatment Interaction (Monotherapy Only)"
subtitle: "Average cell sizes for lapse × DR severity × 4-level treatment across imputations"
author: "Erik Westlund"
date: "`r Sys.Date()`"
format: html
---

@@

```{r setup}
#| include: false
source("dependencies.R")
source("functions.R")  # Load formulas and helper functions
setup_analysis(seed = 2025)

# Additional packages
library(knitr)
library(tidyr)
library(dplyr)

# ANALYSIS MODE: Use same as imputation/PS
analysis_mode <- "final"  # Should match imputation and PS modes

# Set parameters based on mode
if (analysis_mode == "test") {
  max_imputations_to_process <- 2  # Process only 2 imputations in test mode
} else {
  max_imputations_to_process <- NULL  # Process all available imputations
}

# Initialize logging
log_file <- init_log("cell_count_analysis_full_interaction")

# Configuration for cell size checking
MIN_N_PER_CELL <- 5    # Minimum observations per interaction cell

# Treatment approach configuration
TREATMENT_APPROACH <- "hierarchical"  # "hierarchical" or "collapsed"

log_message("======================================")
log_message("Cell Count Analysis - Full Interaction")
log_message("======================================")
log_message(paste0("Analysis started at: ", Sys.time()))
```

## Load Data and Weights

```{r load-data}
# Load imputed datasets
imputed_file <- file.path(reanalysis_data_dir, "imputed_datasets.rds")
if (!file.exists(imputed_file)) {
  stop("Imputed datasets not found. Run reanalysis-2 first.")
}

imputed_datasets <- readRDS(imputed_file)
n_available <- length(imputed_datasets)

# Load PS results for weights
ps_file <- file.path(reanalysis_data_dir, "ps_results.rds")
if (!file.exists(ps_file)) {
  stop("PS results not found. Run reanalysis-4 first.")
}

ps_results <- readRDS(ps_file)
n_ps_available <- length(ps_results$twang_results)

# Determine how many imputations to process
if (!is.null(max_imputations_to_process)) {
  n_imputations <- min(max_imputations_to_process, n_available, n_ps_available)
} else {
  n_imputations <- min(n_available, n_ps_available)
}

log_message(paste0("Available imputed datasets: ", n_available))
log_message(paste0("Available PS weight sets: ", n_ps_available))
log_message(paste0("Processing: ", n_imputations, " imputations"))

# Verify match between imputations and PS weights
n_ps_weights <- length(ps_results$twang_results)
if (n_ps_weights < n_imputations) {
  warning("Only ", n_ps_weights, " PS weight sets available for ", n_imputations, " imputations")
  stop("Mismatch between imputations and PS results. Ensure PS analysis was run with same number of imputations.")
}

for (i in 1:n_imputations) {
  imputed_datasets[[i]]$iptw_weight <- ps_results$twang_results[[i]]$weights
}
```

**Available imputed datasets:** `r n_available`  
**Available PS results:** `r n_ps_available`  
**Processing:** `r n_imputations` **datasets** `r if(analysis_mode == "test") "(test mode limit)" else "(all available)"`

**Note:** This analysis is restricted to monotherapy patients only (n_treatments ≤ 1) to avoid complexity from overlapping treatments.

```{r prepare-data}
# Create consistent analysis variables across all imputed datasets
for (i in 1:n_imputations) {
  imputed_datasets[[i]] <- imputed_datasets[[i]] |>
    mutate(
      # First, check treatment overlaps for diagnostic purposes
      n_treatments = as.numeric(as.character(other_inject)) + 
                     as.numeric(as.character(anti_VEGF)) + 
                     as.numeric(as.character(focal_laser_flag)) + 
                     as.numeric(as.character(PRP_flag)),

      # Create treatment factor based on chosen approach
      treatment_type = case_when(
        TREATMENT_APPROACH == "hierarchical" ~ case_when(
          focal_laser_flag == 1 ~ "focal_laser",  # Highest priority to preserve all 36 cases
          anti_VEGF == 1 ~ "anti_VEGF",
          PRP_flag == 1 ~ "PRP",
          other_inject == 1 ~ "other_inject",
          TRUE ~ "none"
        ),
        TREATMENT_APPROACH == "collapsed" ~ case_when(
          anti_VEGF == 1 ~ "anti_VEGF",
          PRP_flag == 1 ~ "PRP",
          focal_laser_flag == 1 | other_inject == 1 ~ "other_or_focal",
          TRUE ~ "none"
        ),
        TRUE ~ "none"  # fallback
      ),

      # Set factor levels based on approach
      treatment_type = if (TREATMENT_APPROACH == "collapsed") {
        factor(treatment_type, 
               levels = c("none", "other_or_focal", "PRP", "anti_VEGF"))
      } else {
        factor(treatment_type,
               levels = c("none", "other_inject", "focal_laser", "PRP", "anti_VEGF"))
      },

      # Keep binary for comparison
      any_treatment = as.numeric(anti_VEGF == 1 | PRP_flag == 1 | 
                                 other_inject == 1 | focal_laser_flag == 1),
      any_treatment = factor(any_treatment, levels = c(0, 1), 
                            labels = c("No_Treatment", "Any_Treatment")),

@@

      # Ensure lapse is properly coded
      ever_lapse_binary = factor(ever_lapse_binary, levels = c(0, 1),
                                 labels = c("No_Lapse", "Lapsed"))
    ) |>
    # RESTRICT TO MONOTHERAPY PATIENTS (n_treatments == 0 or 1)
    filter(n_treatments <= 1)

  log_message(paste0("Imputation ", i, ": ", nrow(imputed_datasets[[i]]), 
                     " patients after restricting to monotherapy"))
}

# Check treatment distribution and overlaps
first_data <- imputed_datasets[[1]]

log_message(paste0("Using treatment approach: ", TREATMENT_APPROACH))
if (TREATMENT_APPROACH == "collapsed") {
  log_message("Combining focal_laser and other_inject into 'other_or_focal' category")
}

# Check how many patients have multiple treatments
multi_treatment <- table(first_data$n_treatments)
log_message("Number of treatments per patient:")
for (n in names(multi_treatment)) {
  log_message(paste0("  ", n, " treatment(s): ", multi_treatment[n]))
}

# Check specific overlaps with focal laser
if (sum(as.numeric(as.character(first_data$focal_laser_flag))) > 0) {
  focal_patients <- first_data[as.numeric(as.character(first_data$focal_laser_flag)) == 1, ]
  log_message(paste0("Total focal laser patients: ", nrow(focal_patients)))
  log_message(paste0("  Focal only: ", sum(focal_patients$n_treatments == 1)))
  log_message(paste0("  Focal + anti_VEGF: ", sum(as.numeric(as.character(focal_patients$anti_VEGF)))))
  log_message(paste0("  Focal + PRP: ", sum(as.numeric(as.character(focal_patients$PRP_flag)))))
  log_message(paste0("  Focal + other: ", sum(as.numeric(as.character(focal_patients$other_inject)))))
}

treatment_dist <- table(first_data$treatment_type)
log_message("Treatment type distribution (hierarchical assignment):")
for (tx in names(treatment_dist)) {
  log_message(paste0("  ", tx, ": ", treatment_dist[tx]))
}

# Create a cross-tab to show the hierarchical effect
focal_assignment <- table(
  focal_laser_flag = first_data$focal_laser_flag,
  assigned_treatment = first_data$treatment_type
)
log_message("Where focal laser patients were assigned:")
print(focal_assignment)
```

@@

```{r average-cell-counts}
# Compute cell counts for each imputation
all_cell_counts <- list()

@@

kable(average_table,
      caption = paste0("Average cell sizes across ", n_imputations, " imputations"),
      digits = 1,
      col.names = c("Lapse", "DR Severity", names(average_table)[-(1:2)]))

# Identify small cells (averaged across imputations)
small_cells <- average_counts |> 
  filter(mean_n < MIN_N_PER_CELL)

@@

  kable(small_cells |> 
        select(ever_lapse_binary, dr_severity, treatment_type, 
               mean_n, min_n, max_n) |>
        arrange(mean_n),
        caption = paste0("Cells with average count < ", MIN_N_PER_CELL),
        digits = 1,
        col.names = c("Lapse", "DR Severity", "Treatment", "Mean N", "Min N", "Max N"))
}

@@

## Session Info

>>>>

---
title: "Cell Count Analysis for Lapse × DR × Treatment Interaction (Monotherapy Only)"
subtitle: "Checking cell sizes across imputations"
author: "Erik Westlund"
date: "`r Sys.Date()`"
format: html
---

@@

```{r setup}
#| include: false
source("dependencies.R")
source("functions.R")
setup_analysis(seed = 2025)

library(knitr)
library(tidyr)
library(dplyr)

# Initialize logging
log_file <- init_log("cell_count_analysis")

# Configuration
MIN_N_PER_CELL <- 5    # Minimum observations per interaction cell
TREATMENT_APPROACH <- "hierarchical"  # "hierarchical" or "collapsed"

log_message("======================================")
log_message("Cell Count Analysis")
log_message("======================================")
log_message(paste0("Analysis started at: ", Sys.time()))
```

## Load Imputed Datasets

```{r load-data}
# Load imputed datasets from reanalysis-2
imputed_file <- file.path(reanalysis_data_dir, "imputed_datasets.rds")

if (!file.exists(imputed_file)) {
  stop("Imputed datasets not found. Run reanalysis-2-multiple_imputation.qmd first.")
}

imputed_datasets <- readRDS(imputed_file)
n_imputations <- length(imputed_datasets)

log_message(paste0("Loaded ", n_imputations, " imputed datasets"))
```

## Prepare Treatment Variables

```{r prepare-treatment}
# Create treatment variables and restrict to monotherapy
for (i in 1:n_imputations) {
  imputed_datasets[[i]] <- imputed_datasets[[i]] |>
    mutate(
      # Count treatments per patient
      n_treatments = as.numeric(as.character(other_inject)) + 
                     as.numeric(as.character(anti_VEGF)) + 
                     as.numeric(as.character(focal_laser_flag)) + 
                     as.numeric(as.character(PRP_flag)),

      # Create treatment factor with hierarchical assignment
      treatment_type = case_when(
        focal_laser_flag == 1 ~ "focal_laser",  # Highest priority
        anti_VEGF == 1 ~ "anti_VEGF",
        PRP_flag == 1 ~ "PRP",
        other_inject == 1 ~ "other_inject",
        TRUE ~ "none"
      ),

      treatment_type = factor(treatment_type,
                              levels = c("none", "other_inject", "focal_laser", "PRP", "anti_VEGF")),

@@

      # Ensure lapse is properly coded
      ever_lapse_binary = factor(ever_lapse_binary, levels = c(0, 1),
                                 labels = c("No_Lapse", "Lapsed"))
    ) |>
    # RESTRICT TO MONOTHERAPY PATIENTS
    filter(n_treatments <= 1)

  log_message(paste0("Imputation ", i, ": ", nrow(imputed_datasets[[i]]), 
                     " patients after restricting to monotherapy"))
}
```

@@

```{r compute-cell-counts}
# Compute cell counts for each imputation
all_cell_counts <- list()

@@

kable(average_table,
      caption = paste0("Average cell sizes across ", n_imputations, " imputations (monotherapy only)"),
      digits = 1)

# Identify small cells
small_cells <- average_counts |> 
  filter(mean_n < MIN_N_PER_CELL)

@@

  kable(small_cells |> 
        select(ever_lapse_binary, dr_severity, treatment_type, 
               mean_n, min_n, max_n) |>
        arrange(mean_n),
        caption = paste0("Cells with average count < ", MIN_N_PER_CELL),
        digits = 1)
}

@@

## Cell Distribution Details

```{r detailed-distribution}
# Show distribution with standard deviations
detailed_table <- average_counts |>
  mutate(
    mean_sd = paste0(round(mean_n, 1), " (", round(sd_n, 1), ")"),
    range = paste0("[", round(min_n), "-", round(max_n), "]")
  ) |>
  select(ever_lapse_binary, dr_severity, treatment_type, mean_sd, range) |>
  arrange(ever_lapse_binary, dr_severity, treatment_type)

kable(detailed_table,
      caption = "Cell sizes: Mean (SD) and [Min-Max] range across imputations",
      col.names = c("Lapse", "DR Severity", "Treatment", "Mean (SD)", "Range"))
```
